---
title: 'Profiling Youth Sexual Risk in Nigeria: A Machine Learning Approach Using
  Multiple Indicator Cluster Survey (MICS6) Data'
author: "Nasir Umar"
date: "2025-05-15"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

```

This report presents a machine learning analysis of youth sexual risk and media 
engagement patterns in Nigeria using data from the Multiple Indicator Cluster Survey 
(MICS6). Young people aged 15–24 were selected from the full MICS dataset 
and analysed across two thematic tracks:

Track 1: Patterns of access to digital and traditional media platforms 
(e.g., radio, TV, internet, mobile phones).

Track 2: Sexual risk behaviors, focusing on condom use and nature of last sexual partner 
(spousal vs. non-spousal).

We apply unsupervised learning models to identify naturally occurring 
clusters within each track:
K-Means clustering is used for Track 1 (Media Engagement),
Partitioning Around Medoids (PAM) with Gower Distance is used for Track 2 
(Sexual Risk Profiling), to accommodate mixed data types.

We explore how demographic characteristics such as sex and wealth intersect with 
these clusters — and whether digitally engaged youth are more at risk.
```

# Load & Prepare Data 

  - Load needed libraries
    
```{r load needed libraries}
#   Phase 1: Project Setup & Data Loading
library(haven)            # mics is in spss (.sav) file, read raaw data 
#   Phase 2: Data Cleaning, Wrangling, and Preparation
library(tidyverse)        # data cleaning & wrangling
library(data.table)       # data dataframe object handling & manipulation
library(labelled)         # labels & values
library(Hmisc)            # Initial data view - values and label
library(mice)             # to handle missing data, imputation
library(dplyr)            # data prep
library(tidyr)            # data prep
#   Phase 3: Exploratory Data Analysis (EDA) / Descriptives
library(modelsummary)     # descriptive tables
library(kableExtra)       # custom tables
library(sjPlot)           # model visualizations or descriptive plots
library(ggplot2)          # plots, graphs
library(patchwork)        # to combine multiple ggplot2
library(gt)               # tables spec
#   Phase 4: Main Machine Learning Analysis & Model Evaluation
library(tidymodels)       # machine learning 
library(caret)            # machine learning model training 
library(yardstick)        # calculate performance metrics for machine learning models
#   Phase 5: Reporting and Output Generation 
library(flextable)        # custom tables with r markdown
library(knitr)
```

```{r output directories, eval=FALSE}
dir.create("figures", showWarnings = FALSE)
dir.create("tables", showWarnings = FALSE)
dir.create("output", showWarnings = FALSE)
```

# Load, Subset, Clean & Organise Data for Analysis

  - Data Setup & Loading 

```{r load data}
# Data source: MICS6 2023 (Year of interview 2021)
# Data files: 
    # mn.sav (males)
    # wm.sav (females) 

mics6_mn <- read_sav("Data/mn.sav")
mics6_wn <- read_sav("Data/wm.sav")
```

```{r look at the loaded data, eval=FALSE}
view_df(mics6_mn)
glimpse(mics6_mn)
label(mics6_mn)

view_df(mics6_wn)
view(mics6_wn)
```

## Select Var (males) for Analysis
     
```{r subset data male}
mics6_mn_selected <- mics6_mn %>% 
  mutate(
    cluster_id = MWM1,
    household_id = MWM2,
    man_id = MWM3,
    household_uid = paste(cluster_id,household_id, sep = "_"),
    man_uid = paste(cluster_id, household_id, man_id, sep = "_"),
    age_male = MWB4,
    education_ever = MWB5,
    education_level = MWB6A,
    sch_now = MWB9,
    sch_now_level = MWB10A,
    read_can = MWB14,
    state = HH7,
    age_cat = MWAGE,
    newspaper = MMT1,
    radio = MMT2,
    tv = MMT3,
    computer_tab = MMT5, 
    internet = MMT10,
    mphone_own = MMT11,	
    mphone = MMT12,
    sex_last12months = MSB7,
    condom_use = MSB8,
    relations_sexpartner = MSB9,
    urban_rural = HH6,
    education_male = mwelevel,
    survey_weight = mnweight,	
    strata = stratum, 
    wealth_score = wscore,
    wealth_quintile = windex5) %>% 
  select(cluster_id, household_id, man_id, household_uid, 
         man_uid, age_male, education_ever, education_level, 
         sch_now, sch_now_level, read_can, state, age_cat, 
         newspaper, radio, tv, computer_tab,  internet, 
         mphone_own,	mphone, sex_last12months, condom_use, 
         relations_sexpartner, urban_rural, education_male, 
         survey_weight, strata, wealth_score, wealth_quintile)
```

### Data Wrangling Male Subset Dataset
     
```{r clean subset data male}
mn_clean <- mics6_mn_selected %>%
  mutate(
    newspaper = na_if(newspaper, 9),
    radio = na_if(radio, 9),
    tv = na_if(tv, 9),
    computer_tab = na_if(computer_tab, 9),
    internet = na_if(internet, 9),
    mphone = na_if(mphone, 9),
    
    mphone_own = case_when(
      mphone_own == 1 ~ 1,
      mphone_own == 2 ~ 0,
      TRUE ~ NA_real_
    ),
    
    education_ever = na_if(education_ever, 9),
    sch_now = na_if(sch_now, 9),
    sch_now_level = na_if(sch_now_level, 99),
    read_can = na_if(read_can, 9),
    
    urban_rural = case_when(
      urban_rural == 1 ~ "Urban",
      urban_rural == 2 ~ "Rural",
      TRUE ~ NA_character_
    ),
    
    sex_last12months = case_when(
      sex_last12months == 1 ~ 1,
      sex_last12months == 2 ~ 0,
      TRUE ~ NA_real_
    ),
    
    condom_use = case_when(
      condom_use == 1 ~ 0,  
      condom_use == 2 ~ 1,  
      TRUE ~ NA_real_
    ),
    
    relations_sexpartner = case_when(
      relations_sexpartner %in% c(4, 5, 6) ~ 1,  
      relations_sexpartner %in% c(1, 2, 3) ~ 0,  
      TRUE ~ NA_real_
    ),
    
  
    education_male = factor(education_male,
                            levels = c(0, 1, 2, 3, 4, 9),
                            labels = c("None", "Primary", "Jnr Secondary", "Snr Secondary", "Higher", "Missing")),
    
    wealth_quintile = factor(wealth_quintile,
                             levels = 1:5,
                             labels = c("Poorest", "Second", "Middle", "Fourth", "Richest"))
  )

mn_clean <- mn_clean %>%
  filter(age_cat %in% 1:2) %>%
  mutate(sex = "Male")
```

## Select Var (females) for Analysis

```{r subset data female}
mics6_wn_selected <- mics6_wn %>% 
  mutate(
    cluster_id = WM1,
    household_id = WM2,
    woman_id = WM3,
    household_uid = paste(cluster_id,household_id, sep = "_"),
    man_uid = paste(cluster_id, household_id, woman_id, sep = "_"),
    age_female = WB4,
    education_ever = WB5,
    education_level = WB6A,
    sch_now = WB9,
    sch_now_level = WB10A,
    read_can = WB14,
    state = HH7,
    age_cat = WAGE,
    newspaper = MT1,
    radio = MT2,
    tv = MT3,
    computer_tab = MT5, 
    internet = MT10,
    mphone_own = MT11,	
    mphone = MT12,
    sex_last12months = SB7,
    condom_use = SB8,
    relations_sexpartner = SB9,
    urban_rural = HH6,
    education_male = welevel,
    survey_weight = wmweight,	
    strata = stratum, 
    wealth_score = wscore,
    wealth_quintile = windex5) %>% 
  select(cluster_id,  household_id, woman_id, household_uid, 
         man_uid, age_female, education_ever, education_level, 
         sch_now, sch_now_level, read_can, state, age_cat, newspaper, 
         radio,  tv, computer_tab, internet, mphone_own, mphone, 
         sex_last12months, condom_use, relations_sexpartner, 
         urban_rural, education_male, survey_weight, strata, 
         wealth_score, wealth_quintile)
```

### Data Wrangling Female Subset Dataset

```{r clean subset data female}
wn_clean <- mics6_wn_selected %>%
  mutate(sex = "Female") %>%
  
  mutate(
    newspaper = na_if(newspaper, 9),
    radio = na_if(radio, 9),
    tv = na_if(tv, 9),
    computer_tab = na_if(computer_tab, 9),
    internet = na_if(internet, 9),
    mphone = na_if(mphone, 9),
    
    mphone_own = case_when(
      mphone_own == 1 ~ 1,
      mphone_own == 2 ~ 0,
      TRUE ~ NA_real_
    ),
    
    education_ever = na_if(education_ever, 9),
    sch_now = na_if(sch_now, 9),
    sch_now_level = na_if(sch_now_level, 99),
    read_can = na_if(read_can, 9),
    
    urban_rural = case_when(
      urban_rural == 1 ~ "Urban",
      urban_rural == 2 ~ "Rural",
      TRUE ~ NA_character_
    ),
    
    sex_last12months = case_when(
      sex_last12months == 1 ~ 1,
      sex_last12months == 2 ~ 0,
      TRUE ~ NA_real_
    ),
    
    condom_use = case_when(
      condom_use == 1 ~ 0,  
      condom_use == 2 ~ 1,  
      TRUE ~ NA_real_
    ),
    
    relations_sexpartner = case_when(
      relations_sexpartner %in% c(4, 5, 6) ~ 1,  
      relations_sexpartner %in% c(1, 2, 3) ~ 0, 
      TRUE ~ NA_real_
    ),
    
    age_female = as.numeric(age_female),
    
    is_youth = age_cat %in% 1:2  # 1 = 15–19, 2 = 20–24
  ) %>%
  
  filter(is_youth) %>%
  
  mutate(
    education_male = factor(education_male,
                            levels = c(0, 1, 2, 3, 4, 9),
                            labels = c("None", "Primary", "Jnr Secondary", "Snr Secondary", "Higher", "Missing")),
    
    wealth_quintile = factor(wealth_quintile,
                             levels = 1:5,
                             labels = c("Poorest", "Second", "Middle", "Fourth", "Richest"))
  )

```
     
  - Merge Cleaned Male and Female Datasets
     
```{r merge cleaned male and female datasets}
combined_data <- bind_rows(mn_clean, wn_clean)
```

```{r check mergeg cleaned male and female datasets, eval=FALSE}
view_df(combined_data)
glimpse(combined_data)
label(combined_data)
```

# Explore & Visualise Missingness Combined_data Dataset

```{r missingness, eval=FALSE}
library(naniar)
library(dplyr)

miss_var_summary(combined_data)

print(miss_var_summary(combined_data), n = Inf)

combined_data %>%
  select(sex, age_male, age_female, age_cat, newspaper, radio, tv, internet, mphone, 
         sex_last12months, condom_use, relations_sexpartner) %>%
  summarise_all(~ sum(is.na(.)))

vis_miss(combined_data)
```

```{r summarise distribution of missing, eval=FALSE}
combined_data %>%
  summarise(
    age = mean(ifelse(is.na(age_male), age_female, age_male), na.rm = TRUE),
    internet_mean = mean(internet, na.rm = TRUE),
    tv_mean = mean(tv, na.rm = TRUE),
    condom_use_mean = mean(condom_use, na.rm = TRUE)
  )

combined_data %>%
  group_by(sex) %>%
  summarise(
    n = n(),
    prop_newspaper = mean(!is.na(newspaper), na.rm = TRUE),
    sexually_active = mean(sex_last12months == 1, na.rm = TRUE),
    used_condom = mean(condom_use == 0, na.rm = TRUE)
  )
```

  - Filter youth (15–24) 
    
```{r filter yought}
combined_data <- combined_data %>%
  filter(age_cat %in% 1:2) %>%
  mutate(
    age = ifelse(is.na(age_male), age_female, age_male),
    sex = as.character(sex)
  )
```

# Track 1 — Media Clustering (K-Means)

```{r K-Means}
youth_media_data <- combined_data %>%
  select(man_uid, sex, age, newspaper, radio, tv, internet, mphone, mphone_own,
         education_level, wealth_quintile, read_can) %>%
  drop_na()

media_vars_scaled <- youth_media_data %>%
  select(age, newspaper, radio, tv, internet, mphone, mphone_own, education_level, read_can) %>%
  scale()

set.seed(42)
k3 <- kmeans(media_vars_scaled, centers = 3, nstart = 25)
youth_media_data$media_cluster <- factor(k3$cluster)

youth_media_data <- youth_media_data %>%
  mutate(
    media_cluster_label = case_when(
      media_cluster == 1 ~ "Traditional Access",
      media_cluster == 2 ~ "Disconnected Younger Youth",
      media_cluster == 3 ~ "Digital Media Consumers"
    )
  )
```

## PCA visualisation (Track 1)

```{r PCA}
pca_result <- prcomp(media_vars_scaled, center = TRUE, scale. = TRUE)

pca_df_track1 <- as.data.frame(pca_result$x[, 1:2]) %>%
  mutate(media_cluster_label = youth_media_data$media_cluster_label)

cluster_colors <- c(
  "Traditional Access" = "#1f77b4",          
  "Disconnected Younger Youth" = "#ff7f0e",  
  "Digital Media Consumers" = "#2ca02c",     
  "Low-Risk, Low-Media" = "#1f77b4",         
  "Cautious Digital Adopters" = "#ff7f0e",   
  "Digitally Active, High-Risk" = "#2ca02c"  
)

# Plot with ellipses
library(ggplot2)

ggplot(pca_df_track1, aes(x = PC1, y = PC2, color = media_cluster_label)) +
  geom_point(alpha = 0.6, size = 2.5) +
  stat_ellipse(level = 0.95, type = "norm", linetype = "dashed") +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Track 1: Youth Media Engagement",
    x = "Principal Component 1", 
    y = "Principal Component 2", color = "Media Cluster"
  ) +
  theme_minimal()
```

# Track 2 — Sexual Risk Clustering (PAM)

```{r PAM}
track2_data <- combined_data %>%
  filter(sex_last12months == 1) %>%
  select(man_uid, sex, age, condom_use, relations_sexpartner,
         newspaper, radio, tv, internet, mphone, mphone_own,
         education_level, read_can, wealth_quintile) %>%
  drop_na()

track2_clean <- track2_data %>%
  mutate(
    condom_use = as.factor(condom_use),
    relations_sexpartner = as.factor(relations_sexpartner),
    mphone_own = as.factor(mphone_own),
    sex = as.factor(sex),
    wealth_quintile = as.factor(wealth_quintile),
    newspaper = as.numeric(newspaper),
    radio = as.numeric(radio),
    tv = as.numeric(tv),
    internet = as.numeric(internet),
    mphone = as.numeric(mphone),
    education_level = as.numeric(education_level),
    read_can = as.numeric(read_can)
  )

library(cluster)
gower_dist <- daisy(track2_clean %>% select(-man_uid, -sex, -age), metric = "gower")
pam_fit <- pam(gower_dist, k = 3)

track2_data$risk_cluster <- as.factor(pam_fit$clustering)

track2_data <- track2_data %>%
  mutate(
    risk_cluster_label = case_when(
      risk_cluster == "1" ~ "Low-Risk, Low-Media",
      risk_cluster == "2" ~ "Cautious Digital Adopters",
      risk_cluster == "3" ~ "Digitally Active, High-Risk"
    )
  )
```

## Visualise PAM Clusters (Track 2) 

```{r}
track2_pca_data <- track2_data %>%
  select(newspaper, radio, tv, internet, mphone, education_level, read_can)

track2_pca_scaled <- scale(track2_pca_data)
pca_track2 <- prcomp(track2_pca_scaled)

pca_df_track2 <- as.data.frame(pca_track2$x[, 1:2])
pca_df_track2$risk_cluster_label <- track2_data$risk_cluster_label

cluster_colors <- c(
  "Low-Risk, Low-Media" = "#1b9e77",
  "Cautious Digital Adopters" = "#d95f02",
  "Digitally Active, High-Risk" = "#7570b3"
)

# Plot with ellipses
library(ggplot2)

ggplot(pca_df_track2, aes(x = PC1, y = PC2, color = risk_cluster_label)) +
  geom_point(size = 2.5, alpha = 0.7) +
  stat_ellipse(level = 0.95, linetype = "dashed") +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Track 2: Youth Sexual Risk Profiles (PCA)",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Risk Cluster"
  ) +
  theme_minimal()
```

    - Merge Tracks
    
```{r merge tracks on "man_uid"}
track1_data <- youth_media_data %>%
  select(man_uid, media_cluster_label)

track1_vs_2 <- left_join(track2_data, track1_data, by = "man_uid")
```

# Comparison of Clusters (media & risk)

## Side-by-Side PCA Cluster Visualisation

```{r r track1-track2-pca-combined, fig.width=14, fig.height=8.5, out.width='100%', fig.cap="Figure: Side-by-Side PCA Plots of Youth Media (Track 1) and Sexual Risk (Track 2) Clusters", message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)

cluster_colors <- c(
  "Traditional Access" = "#1f77b4",          
  "Disconnected Younger Youth" = "#ff7f0e",  
  "Digital Media Consumers" = "#2ca02c",     
  "Low-Risk, Low-Media" = "#1f77b4",         
  "Cautious Digital Adopters" = "#ff7f0e",   
  "Digitally Active, High-Risk" = "#2ca02c"  
)

pca_df_track1 <- as.data.frame(pca_result$x[, 1:2]) %>%
  mutate(cluster = youth_media_data$media_cluster_label)

p1 <- ggplot(pca_df_track1, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.6, size = 2.5) +
  stat_ellipse(level = 0.95, type = "norm", linetype = "dashed") +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Track 1: Youth Media Engagement",
    x = "PC1", y = "PC2", color = "Media Cluster"
  ) +
  theme_minimal()

p2 <- ggplot(pca_df_track2, aes(x = PC1, y = PC2, color = risk_cluster_label)) +
  geom_point(alpha = 0.6, size = 2.5) +
  stat_ellipse(level = 0.95, type = "norm", linetype = "dashed") +
  scale_color_manual(values = cluster_colors) +
  labs(
    title = "Track 2: Youth Sexual Risk Profiles",
    x = "PC1", y = "PC2", color = "Risk Cluster"
  ) +
  theme_minimal()

combined_plot <- p1 + p2 + 
  plot_layout(ncol = 2, guides = "collect") & 
  theme(legend.position = "bottom")

combined_plot
```

  - Export Plot (PNG and PDF)
     
```{r export plots}
ggsave(
  filename = "figures/track1_track2_side_by_side.pdf",
  plot = combined_plot,
  width = 14,
  height = 7
)

ggsave(
  filename = "figures/track1_track2_side_by_side.png",
  plot = combined_plot,
  width = 14,
  height = 7,
  dpi = 300
)
```

# Clusters by Demographics

## Track 1 – Media Cluster by sex & wealth
    
```{r media sex & wealth}
track1_summary <- youth_media_data %>%
  group_by(media_cluster_label, sex, wealth_quintile) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(media_cluster_label) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup()

track1_table <- track1_summary %>%
  unite("Group", sex, wealth_quintile, sep = " - ") %>%
  select(media_cluster_label, Group, pct) %>%
  tidyr::pivot_wider(names_from = Group, values_from = pct)

track1_table %>%
  gt() %>%
  tab_header(
    title = "Demographic Distribution by Media Cluster (Track 1)",
    subtitle = "Percentage of respondents by sex and wealth within each media cluster"
  ) %>%
  fmt_percent(columns = everything(), scale_values = FALSE)
```

## Track 2 – Risk Cluster by sex & wealth

```{r risk sex & wealth}
track2_summary <- track2_data %>%
  group_by(risk_cluster_label, sex, wealth_quintile) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(risk_cluster_label) %>%
  mutate(pct = round(100 * n / sum(n), 1)) %>%
  ungroup()

track2_table <- track2_summary %>%
  unite("Group", sex, wealth_quintile, sep = " - ") %>%
  select(risk_cluster_label, Group, pct) %>%
  tidyr::pivot_wider(names_from = Group, values_from = pct)

track2_table %>%
  kbl(caption = "Demographic Distribution by Sexual Risk Cluster (Track 2)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) 
  
```


```
Discussion and Implications

This analysis revealed three distinct media engagement clusters and three 
sexual risk profiles among Nigerian youth:

Digital Media Consumers tend to have higher access to internet and mobile phones, 
and show some overlap with higher-risk sexual behavior.

Disconnected Younger Youth show low digital engagement and tend to be less sexually 
active or lower-risk.

Cautious Digital Adopters and Low-Risk, Low-Media profiles suggest that access 
alone does not equate to risk.

The integration of Track 1 and Track 2 clusters reveals important overlaps — but also 
highlights that digital engagement alone does not uniformly predict risk. Instead, 
demographic context (such as sex, wealth) remains crucial.

Policy Relevance

Targeted Sexual Health Interventions: Programs may need to differentiate between 
digitally connected yet cautious youth versus those who are digitally active 
and high-risk.

Digital Literacy and Safe Behavior Campaigns:** Especially for those with high 
internet/mobile access.

Addressing Inequity: Disconnected youth may lack both sexual health information 
and general digital opportunities.

Limitations

Survey weights were not applied; hence findings may not be nationally representative.
Unmeasured variables (e.g., peer influence, religious beliefs) may affect behaviors.

Next Steps

- Extend to rural vs. urban comparisons.
- Include additional behavioral indicators if available.
- Apply supervised machine learning for prediction (e.g., classification models).

```




